---
title: "Modeling High Risk PAID Scores of Participants in Diabetes Self-Management Program"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

## Project Motivation

Type 2 Diabetes is a chronic metabolic disorder that affects many
Americans characterized by high blood glucose levels through insulin
resistance. It can lead to complications elsewhere in the body, such as
retinopathies, kidney failure, and amputation of limbs due to poor wound
healing. Within the United States, T2 Diabetes affects 26 million
people. ([NIH T2
Diabetes](https://www.nih.gov/research-training/accelerating-medicines-partnership-amp/type-2-diabetes)).

As of 2017, the annual burden of diabetes treatment on the US healthcare
system is approximately 327 billion USD. Of this approximation, 237
billion USD is associated with direct medical costs, including inpatient
hospital treatment, prescriptions and medical supplies, and outpatient
visits. ([American Diabetes Association
Stats](https://www.diabetes.org/resources/statistics/cost-diabetes)).

The Problem Areas in Diabetes (PAID) questionnaire is a short,
self-guided questionnaire completed by patients with diabetes. The PAID
questionnaire is a robust measure that can evaluate emotional distress
as well as predict future blood glucose control in patients with
diabetes. The questionnaire is scored on a scale from 0-100; scores of
40 and higher are strongly associated with emotional burnout, while
scores below 10 with poor glucose control are associated with patients
in denial of their condition. ([PAID Questionnaire
UConn](https://www.huskyhealthct.org/providers/provider_postings/diabetes/PAID_problem_areas_in_diabetes_questionnaire.pdf)).

```{r Setup, message=FALSE, include=FALSE}
library(tidyverse)
library(fastDummies)
library(caret)
library(magrittr)
library(MLeval)
library(microbenchmark)
library(tictoc)

set.seed(17)

diabetes_raw <- read_csv(
  "diabetes_self_management.csv",
  col_names = c("class", "class_lang","age", "year","gender", "insurance", "medical_home","ethnicity","education_max", "diabetes", "heart_disease","high_blood_pressure","tobacco_use","previous_diabetes_education","diabetes_knowledge","fruit_veg_freq_daily", "sugary_bev_freq_daily", "food_measurement_freq_weekly", "carb_counting_freq_weekly", "exercise_freq_weekly","PAID_score"),
  col_types = list(
    age = col_integer(),
    year = col_integer(),
    PAID_score = col_double(),
    diabetes_knowledge = col_factor(levels = c("Poor", "Fair", "Good"))
    ),
  skip = 1
  )
```

## Exploratory Data Analysis

```{r Cleaning, include=FALSE}
diabetes_recoded <- diabetes_raw %>%
  mutate(gender = fct_recode(gender,"F" = "f"),
         gender = as.character(gender),
         insurance = str_replace_all(insurance, "MediCARE", "MC"),
         insurance = str_replace_all(insurance, "Medicaid", "MA"),
         insurance = str_replace_all(insurance, "Private insurance", "Private"),
         ethnicity = na_if(ethnicity, "Unknown"),
         diabetes = na_if(diabetes, "Unknown"),
         medical_home = fct_recode(medical_home, "No regular place of care" = "Emergency Room"),
         fruit_veg_freq_daily = na_if(fruit_veg_freq_daily, "Not Sure"),
         fruit_veg_freq_daily = fct_relevel(fruit_veg_freq_daily, levels = c("0","1-2","3-4","5 or more")),
         sugary_bev_freq_daily = na_if(sugary_bev_freq_daily, "Not Sure"),
         sugary_bev_freq_daily = fct_relevel(sugary_bev_freq_daily, levels = c("0","1","2","3 or more")),
         food_measurement_freq_weekly = na_if(food_measurement_freq_weekly, "Not Sure"),
         food_measurement_freq_weekly = fct_collapse(food_measurement_freq_weekly, "0" = c("I don't know how", "0 days")),
         food_measurement_freq_weekly = fct_relevel(food_measurement_freq_weekly,levels = c("0","1-3","4 or more")),
         carb_counting_freq_weekly = na_if(carb_counting_freq_weekly, "Not Sure"),
         carb_counting_freq_weekly = fct_collapse(carb_counting_freq_weekly, "0" = c("I don't know how", "0 days")),
         carb_counting_freq_weekly = fct_relevel(carb_counting_freq_weekly,levels = c("0", "1-3","4 or more")),
         exercise_freq_weekly = na_if(exercise_freq_weekly, "Not Sure"),
         exercise_freq_weekly = str_replace_all(exercise_freq_weekly, "days",""),
         exercise_freq_weekly = str_replace_all(exercise_freq_weekly, "day",""),
         exercise_freq_weekly = fct_relevel(exercise_freq_weekly, levels = c("0 ", "1", "2", "3", "4", "5 or more"))
  ) %>%
  mutate(low_income = ifelse(insurance == "MA" | insurance == "MAP", "Yes", "No"),
         PAID_high_risk = ifelse(PAID_score >= 40, "Yes", "No")
         ) %>%
  relocate(low_income, .after = insurance) %>%
  mutate(
    has_pcp = ifelse(medical_home == "No regular place of care", "No", "Yes")) %>%
  relocate(has_pcp, .after = medical_home)
```

```{r Data Dictionary, echo=FALSE}

data_dictionary <- tribble(
  ~Variable, ~Definition, ~Type,
  #--|----|----
  "class","Diabetes self-management class code", "character",
  "class_lang", "Language of diabetes self-management class", "character",
  "age", "Age of participant", "numeric",
  "year", "Year that survey was completed", "numeric",
  "gender", "Gender of participant", "character",
  "insurance", "Type of insurance held by participant", "character",
  "medical_home", "Place of primary/regular medical care", "character",
  "ethnicity", "Ethnicity of participant", "character",
  "education_max", "Highest completed education level of participant", "character",
  "diabetes", "Does the participant have diabetes?", "character",
  "heart_disease", "Does the participant have heart disease?", "character",
  "tobacco_use", "Does the participant use tobacco products?", "character",
  "previous_diabetes_education", "Has the participant received diabetes education in the past?", "character",
  "diabetes_knowledge", "level of participant's knowledge on diabetes", "ordered factor",
  "fruit_veg_daily", "number of fruits/vegetables consumed daily", "character",
  "sugary_bev_daily", "number of sugary beverages consumed daily", "character",
  "food_measurement_weekly", "number of days per week that participant measures their food consumption", "character",
  "carb_counting_weekly", "number of days per week that participant counts their carbohydrate consumption", "character",
  "exercise_freq_weekly", "number of days per week that participant exercises aerobically", "character",
  "PAID_score", "Problem Areas in Diabetes score from PAID self-questionnaire", "numeric",
  "PAID_high_risk", "Is the PAID score considered to be high risk (equal or greater than 40)?", "character",
  "has_pcp", "Does the participant have a primary care medical home?", "character",
  "low_income", "Does the participant qualify for insurance for low income individuals?","character"
  )

```

### Dataset description

The data for this project was obtained through the [Datacamp Career Hub
Repository](https://github.com/datacamp/careerhub-data/tree/master/Diabetes%20Self-Management).
It is publicly sourced from Public Health Department of Austin, Texas
through their Diabetes Self-Management Education program. Each
observation from the data set are responses from individual participants
who completed the program.

The raw dataset contains 21 unique variables, 3 of which are numeric and
18 that are categorical. The majority of the variables appear to be
self-explanatory, while some are open to interpretation. I have defined
the variables in a data dictionary below, as one was not provided with
the dataset. For the purposes of my analysis, I have created three
additional categorical variables, which I have also defined in the data
dictionary. 'has_pcp' is a simplification of the 'medical_home'
variable. It is not particularly important which specific clinic that a
participant goes for their primary care, but rather that they *do* have
a primary care provider with whom they can follow up regularly over
time. Emergency rooms are acute care providers, and not primary care
providers.

'low_income' is an indicator variable which describes if the participant
has a low income, dependent on their 'insurance' enrollment.
Participants must have a recognized low gross income to qualify for
either Medicaid or MAP (Medical Access Program).

'PAID_high_risk' is an indicator variable dependent on the 'PAID_score'
of the participant. Without any data to assess a participant's glucose
control, it is difficult to assess whether a participant is in denial of
their condition. A 'PAID_score' \>= 40 is considered to be high risk, as
it is shown to indicate emotional burnout in a participant. For this
analysis, 'PAID_high_risk' will be the response variable for
classification.

There are 1688 observations in this dataset.

However, across the dataset, there are many variables with missing
values. Some of these values are missing not at random and can be
explained. Cases where the participant does not have diabetes will not
have a PAID score, and can not have their risk evaluated. Similarly,
variables such as 'has_pcp' and 'low_income' are dependent on
'medical_home' and 'insurance', respectively, and will therefore have
the same missing values as their parent columns.

However, the remaining missing values are of unknown nature. The raw
dataset did not provide any guidance on the interpretation of these
missing values. Dropping observations with any missing values will
deplete the dataset drastically. One-hot encoding the variables will be the better
option to pursue here, including "variable_NA" columns to possibly
identify any other variables with missing not at random values.

```{r dimensions, message=FALSE}
#dimensions of data frame
dim(diabetes_recoded)

#missing values for each column
missing_values <- diabetes_recoded %>% 
  select(everything()) %>%
  summarize_all(funs(sum(is.na(.))))


```

### Exploratory Data Analysis

```{r EDA}
#diabetes status distribution
diabetes_recoded %>% 
  ggplot(aes(x = diabetes)) + geom_bar()

#high-risk patients
diabetes_recoded %>% 
  ggplot(aes(x = PAID_high_risk, fill = diabetes)) + geom_bar(position = "dodge")

#medical_home distribution
diabetes_recoded %>%
  ggplot(aes(x = medical_home, fill = diabetes)) + geom_bar() + coord_flip()

diabetes_recoded %>%
  ggplot(aes(x = has_pcp)) + geom_bar(position = "dodge") + facet_wrap(~ diabetes)

#insurance distribution
diabetes_recoded %>%
  ggplot(aes(x = insurance)) + geom_bar() + coord_flip()

diabetes_recoded %>%
  ggplot(aes(x = low_income, fill = diabetes)) + geom_bar(position = "dodge")

#class distribution
diabetes_recoded %>%
  ggplot(aes(x = diabetes, fill = class)) + geom_bar(position = "dodge")

#class_lang distribution
diabetes_recoded %>%
  ggplot(aes(x = diabetes, fill = class_lang)) + geom_bar(position = "dodge")

#age distribution
diabetes_recoded %>%
  filter(!is.na(gender), !is.na(age)) %>%
  ggplot(aes(x = age, color = gender)) + geom_bar() +
  facet_wrap(~ diabetes)

#ethnicity distribution
diabetes_recoded %>%
  ggplot(aes(x = ethnicity, fill = diabetes)) + geom_bar(position = "dodge") +
  coord_flip()

```

```{r}


#valid PAID scores
diabetes_recoded %>% 
  filter(!is.na(PAID_score)) %>%
  count()

diabetes_recoded %>%
  ggplot(aes(x = PAID_score, fill = PAID_high_risk)) + geom_bar() +
  geom_vline(xintercept = 40)



#age distribution by gender of diabetes patients
diabetes_recoded %>% 
  filter(diabetes == "Yes", !is.na(gender)) %>%
  ggplot(aes(x = gender, y = age)) + geom_boxplot() + coord_flip()


```

## Model Development

```{r Dataset Treatment}
#creating separate data frame for response variable
diabetes_treated_y <- diabetes_recoded %>%
  select(PAID_high_risk)

#one-hot encoding character variables
charvars <- c("ethnicity", "diabetes", "heart_disease", "high_blood_pressure", "tobacco_use", "previous_diabetes_education", "insurance", "gender", "has_pcp", "low_income")

characters <- diabetes_recoded %>%
  select(all_of(charvars)) %>%
  dummy_cols() %>%
  select(-all_of(charvars))

#one-hot encoding ordinal variables
ordvars <- c("education_max", "diabetes_knowledge", "fruit_veg_freq_daily", "sugary_bev_freq_daily", "food_measurement_freq_weekly", "carb_counting_freq_weekly", "exercise_freq_weekly")
level <- c("0","1","2")

ordinals <- diabetes_recoded %>%
  select(all_of(ordvars)) %>%
  mutate(
    education_max = fct_collapse(education_max,
                                 "0" = c("1-8", "9-11", "n", "none", "y"),
                                 "1" = c("Some College", "High School GED"),
                                 "2" = "College"
                                 ),
    education_max = fct_relevel(education_max, levels = level),
    diabetes_knowledge = fct_recode(diabetes_knowledge,
                                    "0" = "Poor",
                                    "1" = "Fair",
                                    "2" = "Good"
                                    ),
    diabetes_knowledge = fct_relevel(diabetes_knowledge, levels = level),
    fruit_veg_freq_daily = fct_collapse(fruit_veg_freq_daily,
                                      "0" = c("0", "1-2"),
                                      "1" = "3-4",
                                      "2" = "5 or more"
                                      ),
    fruit_veg_freq_daily = fct_relevel(fruit_veg_freq_daily, levels = level),
    sugary_bev_freq_daily = fct_collapse(sugary_bev_freq_daily,
                                       "0" = "0",
                                       "1" = c("1","2"),
                                       "2" = "3 or more"
                                       ),
    sugary_bev_freq_daily = fct_relevel(sugary_bev_freq_daily, levels = level),
    food_measurement_freq_weekly = fct_recode(food_measurement_freq_weekly,
                                              "0" = "0",
                                              "1" = "1-3",
                                              "2" = "4 or more"
                                              ),
    food_measurement_freq_weekly = fct_relevel(food_measurement_freq_weekly, levels = level),
    carb_counting_freq_weekly = fct_recode(carb_counting_freq_weekly,
                                            "0" = "0",
                                            "1" = "1-3",
                                            "2" = "4 or more"
                                           ),
    carb_counting_freq_weekly = fct_relevel(carb_counting_freq_weekly, levels = level),
    exercise_freq_weekly = str_trim(exercise_freq_weekly),
    exercise_freq_weekly = fct_recode(exercise_freq_weekly,
                                      "0" = "0",
                                      "0" = "1",
                                      "0" = "2",
                                      "1" = "3", 
                                      "1" = "4",
                                      "2" = "5 or more"
                                      ),
    exercise_freq_weekly = fct_relevel(exercise_freq_weekly, levels = level)
  ) %>%
  dummy_cols() %>%
  select(-all_of(ordvars))

#bringing the tables together
diabetes_treated_x <- diabetes_recoded %>%
  select(-class, -class_lang, -medical_home, -PAID_score, -year) %>%
  select(-all_of(ordvars), -all_of(charvars), -PAID_high_risk) %>%
  bind_cols(characters, ordinals) 

#converting all NAs to zeros (NAs are counted in separate columns)
diabetes_treated_x[is.na(diabetes_treated_x)] <- 0

#validating that there are no missing values
missing_values_treated <- diabetes_treated_x %>% 
  select(everything()) %>%
  summarize_all(funs(sum(is.na(.))))

```

Prior to modeling the data, the dataset will have to be treated.As discussed earlier, the variables in the dataset will be one-hot encoded to address the many missing values in the dataset without dropping observations.The dataset will be split into two tables: one containing predictor variables and the other containing only the response variable, PAID_high_risk. The table of predictor variables needed to be modified to conduct logistic regression. 

To do so, the predictor variables were grouped by character variables, which consist of survey questions indicating status with Y/N/NA answers (e.g. whether the participant has diabetes) and ordinal variables, which consist of survey questions with answers on a spectrum (e.g. how many days in a week does the participant exercise). For the purposes of logistic regression modeling and to reduce the number of dummy columns, the answers to the ordinal variables were releveled to follow "low, medium, high" scoring. The character and ordinal variable tables were then converted so that each variable had a column for each possible response, including NA values. The characters and ordinals tables were joined together to make a treated table of all predictors.



```{r Modeling Prep}
#creating training set and test set

diabetes_modeling_set <- bind_cols(diabetes_treated_x, diabetes_treated_y) %>%
  filter(age != 0) %>%
  mutate(PAID_high_risk = as_factor(PAID_high_risk))

#real world test sets
diabetes_rw <- diabetes_modeling_set %>%
  filter(diabetes_No > 0 | diabetes_NA > 0) %>%
  select(-diabetes_No, -diabetes_Yes, -diabetes_NA)

#training sets
diabetes_train <- diabetes_modeling_set %>%
  filter(diabetes_Yes > 0) %>%
  select(-diabetes_No, -diabetes_Yes, -diabetes_NA) %>%
  filter(PAID_high_risk %in% c("Yes","No"))

# diabetes_train_x <- diabetes_train %>%
#   select(-PAID_high_risk)
 
# diabetes_train_y <- diabetes_train %>%
#   select(PAID_high_risk)
  
control <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  classProbs = TRUE,
  verboseIter = FALSE
)
```

The treated predictors table was joined with the response table. The training dataset consisted of all participants with diabetes and a valid PAID score. The test dataset comprises all participants without diabetes or do not have a valid status.

Since the training dataset only contains 621 observations, each model will be trained using repeated cross validation 10 times with 5 repeats.yes

```{r Model Generation and Evaluation}
#generating and evaluating multiple models

#benchmarked top 4 models

mbm2 <- microbenchmark(
  "glm_zv_knn" = {glm_model_zv_knn <- train(
  PAID_high_risk ~ .,
  data = diabetes_train,
  metric = "ROC",
  method = "glm",
  trControl = control,
  preProcess = c("zv", "knnImpute", "center", "scale"),
  na.action = na.pass
)},
  "glmnet" = {glmnet_model <- train(
  PAID_high_risk ~ .,
  data = diabetes_train,
  metric = "ROC",
  method = "glmnet",
  trControl = control,
  preProcess = c("medianImpute", "center", "scale"),
  na.action = na.pass
)},
  "rf_model_zv_knn" = {rf_model_zv_knn <- train(
  PAID_high_risk ~ .,
  data = diabetes_train,
  metric = "ROC",
  method = "ranger",
  trControl = control,
  preProcess = c("zv" ,"knnImpute", "center", "scale"),
  na.action = na.pass,
  importance = 'impurity'
)},
  "gbm_model" = {gbm_model <- train(
  PAID_high_risk ~ .,
  data = diabetes_train,
  metric = "ROC",
  method = "gbm",
  train.fraction = 0.5,
  trControl = control
)},
  times = 10)


#glm models
# glm_model_median <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "glm",
#   trControl = control,
#   preProcess = c("medianImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7182

# glm_model_nzv_median <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "glm",
#   trControl = control,
#   preProcess = c("nzv", "medianImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7099

# glm_model_zv_median <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "glm",
#   trControl = control,
#   preProcess = c("zv", "medianImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7146

# glm_model_knn <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "glm",
#   trControl = control,
#   preProcess = c("knnImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7156

# glm_model_nzv_knn <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "glm",
#   trControl = control,
#   preProcess = c("nzv", "knnImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7107

# glm_model_zv_knn <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "glm",
#   trControl = control,
#   preProcess = c("zv", "knnImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7206

#glmnet
# glmnet_model <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "glmnet",
#   trControl = control,
#   preProcess = c("medianImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7226246, alpha = 0.1,  lambda = 0.02708264

#random forest models
# rf_model_median <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("medianImpute", "center", "scale"),
#   na.action = na.pass,
#   importance = 'impurity'
# )
#ROC = 0.7366

# rf_model_nzv_median <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("nzv","medianImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7377

# rf_model_zv_median <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("zv","medianImpute", "center", "scale"),
#   na.action = na.pass,
#   importance = 'impurity'
# )
#ROC = 0.7391

# rf_model_pca_median <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("zv","medianImpute", "center", "scale", "pca"),
#   na.action = na.pass
# )
#ROC = 0.7201

# rf_model_knn <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("knnImpute", "center", "scale"),
#   na.action = na.pass
# )
#ROC = 0.7411

# rf_model_nzv_knn <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("nzv", "knnImpute", "center", "scale"),
#   na.action = na.pass,
#   importance = 'impurity'
# )
#ROC = 0.7352

# rf_model_zv_knn <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("zv" ,"knnImpute", "center", "scale"),
#   na.action = na.pass,
#   importance = 'impurity'
# )
#ROC = 0.7427

# rf_model_pca_knn <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "ranger",
#   trControl = control,
#   preProcess = c("zv","knnImpute", "center", "scale", "pca"),
#   na.action = na.pass
# )
#ROC = 0.7141

#gradient boosting
# gbm_model <- train(
#   PAID_high_risk ~ .,
#   data = diabetes_train,
#   metric = "ROC",
#   method = "gbm",
#   train.fraction = 0.5,
#   trControl = control
# )
#ROC = 0.7209 

# print(varImp(rf_model_zv_knn))

```

```{r Model Implementation & Real World Validation, eval=FALSE, include=FALSE}

#creating prediction output table
diabetes_rw_pred <- predict(rf_model_zv_knn, newdata = diabetes_rw, type = "prob")
diabetes_rw_pred <- diabetes_rw_pred %>%
  mutate(prediction = ifelse(Yes > 0.5, "Yes", "No"))

#separate table for final counts
PAID_high_risk_pred <- diabetes_rw_pred %>%
  group_by(prediction) %>%
  summarize(total = n())


```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or
by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output
will be saved alongside it (click the *Preview* button or press
*Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the
editor. Consequently, unlike *Knit*, *Preview* does not run any R code
chunks. Instead, the output of the chunk when it was last run in the
editor is displayed.
